Complementary roles of Theta and Alpha in Multi-Item Associations

### 2019-06-05 manuscript.tex
updated supplemental deep leabra fig and text

The `Deep LEABRA' framework currently being developed in the Emergent software is the initial explorations into the modeling of interactions between thalamic relay nuclei, and superficial vs. deep cortical layers \cite{OReillyWyatteRohrlich14}.  This framework proposes that the neocortex is composed of two separable networks: superficial and deep/thalamic. The superficial-layer network consists of neocortical layers 4, 2, and 3, across different brain areas.  The deep/thalamic network starts in each area with the layer 5b intrinsic bursting (IB) neurons, which receive inputs from local superficial neurons and top-down projections from other areas, and provide strong driving feedforward input to higher-order thalamic nuclei. These 5IB neurons also project to deep layer 6 cortical-thalmic (6CT) neurons, which interconnects with the thalamus (which in turn projects back up to layer 4 of the superficial network and layer 6 in the deep network).  This connectivity can be seen in Figure \ref{fig:attn_compute}.  The current work takes advantage of the biological referents and computational implementation of the Deep LEABRA framework.  However, here it is used as a simple form of attentional modulation (and not a temporal learning mechanism), and the minimal description of the components necessary for this implementation are described below.

The modulation of standard LEABRA unit activations (referred to as superficial units here) based the Deep LEABRA framework and the thalamic gating model described in \textcite{KetzJensenOReilly15} is carried out in two basic steps.  First, a Thalamic Relay Cell (TRC) unit receives input based on the threshold filtered output of some upstream 5IB units (which have received its input from superficial units  processes a given stimulus).  These sending units, for a given TRC, are assumed to be the cortical thalamic projections from neocortical layers 5IB, 6CT \cite{ShermanGuillery06}, which reflect a thresholded filtering of superficial layer activity from both top-down and bottom-up sources. TRC activation dynamics are calculated in a similar manner to all LEABRA units, described in detail in \cite{KetzMorkondaOReilly13}, with the exception that its inputs are subject to an activation threshold of 0.1, and the weighted connections into and out of the TRC are not learned but instead remain fixed at 0.8.

@@

\begin{figure}
  \centering\includegraphics[width=.75\textwidth]{figs/fig_deepleabra_attn_compute.pdf}
  \caption[Signal Flow Through Cortical Layers in Deep LEABRA]{\small How attentional modulation is computed across the deep layers in response to a top-down attentional focus.  Layer 4 receives bottom-up sensory input (initially equally weighted), which then drives superficial layers (2/3), which initially do not reflect the attentional modulation (not shown).  The deep 5IB neurons integrate deep-to-deep top-down attentional inputs plus the local stimulus features from 2/3, to produce the {\em raw} deep output, prior to the contextual normalization process.  6CT then integrates this direct activation from 5IB, to produce, for the first time in the circuit, a properly renormalized multiplicative gain-field activation pattern, with surround inhibition both within the 6CT layer and further downstream in the  TRC circuit providing the renormalization process.  These 6CT activations then modulate (multiply) the superficial-layer activations to produce {\em both} an increase in the attended location, and a decrease for the unattended location, as shown.  In the biology, this modulation affects the layer 4 inputs (not shown) as well as 2/3.  Our model subsumes layer 4 into layer 2/3 neurons.}
  \label{fig:attn_compute}
\end{figure}

>>>>

The `Deep LEABRA' framework currently being developed in the Emergent software is the initial explorations into the modeling of interactions between thalamic relay nuclei, and superficial vs. deep cortical layers \cite{OReillyWyatteRohrlich14}.  This framework proposes that the neocortex is composed of two separable networks: superficial and deep/thalamic. The superficial-layer network consists of neocortical layers 4, 2, and 3, across different brain areas.  The deep/thalamic network starts in each cortical area with the layer 5b intrinsic bursting (IB) neurons, which receive inputs from local superficial neurons and top-down projections from other areas, and provide strong driving feedforward input to higher-order thalamic nuclei. These 5IB neurons also project to deep layer 6 cortical-thalmic (6CT) neurons, which interconnects with the thalamus (which in turn projects back up to layer 4 of the superficial network and layer 6 in the deep network).  This connectivity is illustrated in Figure \ref{fig:attn_compute}.  The current work takes advantage of the biological referents and computational implementation of the Deep LEABRA frameworkm however, here it is used as a simple form of attentional modulation (and not a temporal learning mechanism as described in \textcite{OReillyWyatteRohrlich14}).  The minimal description of the components necessary for this implementation are described below.

The modulation of standard LEABRA unit activations (referred to as superficial units here) based the Deep LEABRA framework and the thalamic gating model described in \textcite{KetzJensenOReilly15} is carried out in two basic steps.  First, a Thalamic Relay Cell (TRC) unit receives input based on the threshold filtered output of some upstream units which have received its input from processing a given stimulus (in this case the hippocampus is providing this top-down modulation based on what is being retrieved).  These sending units, for a given TRC, are assumed to be the cortical thalamic projections from neocortical layers 5IB, 6CT \cite{ShermanGuillery06}, which reflect a thresholded filtering of superficial layer activity from both top-down and bottom-up sources. TRC activation dynamics are calculated in a similar manner to all LEABRA units, described in detail in \cite{KetzMorkondaOReilly13}, with the exception that its inputs are subject to an activation threshold of 0.1, and the weighted connections into and out of the TRC are not learned but instead remain fixed at 0.8.

@@

\begin{figure}
  \centering\includegraphics[width=.75\textwidth]{figs/fig_deepleabra_attn_compute.pdf}
  \caption[Signal Flow Through Cortical Layers in Deep LEABRA]{\small How attentional modulation is computed in response to a top-down attentional focus.  Layer 4 receives bottom-up sensory input, which then drives superficial layers (2/3), which initially do not reflect the attentional modulation (not shown).  The deep layer 5 Intrinsic Bursting (5IB) neurons integrate deep-to-deep top-down attentional inputs plus the local stimulus features from 2/3, to produce a {\em raw} deep output, prior to the contextual normalization process.  Layer 6 Cortical-Thalamic (6CT) neurons then integrate this direct activation from 5IB, to produce, for the first time in the circuit, a properly renormalized multiplicative gain-field activation pattern, with surround inhibition both within the 6CT layer and further downstream in the  TRC circuit providing the renormalization process.  These 6CT activations then modulate (multiply) the superficial-layer activations to produce {\em both} an increase in the attended location, and a decrease for the unattended location, as shown.  
  %In biological systems, this modulation also affects the layer 4 inputs (not shown) as well as 2/3, however, our model subsumes layer 4 into layer 2/3 neurons.
  }
  \label{fig:attn_compute}
\end{figure}
